Handling String-Number Conversion in String Constraint Solving

### 2019-11-07 draft.tex
polish str-int conv encoding

\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\sti}[1]{\mbox{\textbf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textbf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{\mbox{$V^1_{\Sigma_\epsilon}$}}
\newcommand{\cvartwo}{\mbox{$V^2_{\Sigma_\epsilon}$}}
\newcommand{\modelof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}
\newcommand{\parikhof}[1]{\mathit{Parikh}{(#1)}}
\newcommand{\parikhwof}[2]{|#1|_{#2}}

@@

We use $\mathbb{N}$ and $\mathbb{Z}$ to denote the sets of natural numbers (including 0) and 
integers. For a set $A$, we use $|A|$ to denote its size. 
An \emph{alphabet} is a finite set $\Sigma$ of \emph{characters} and a \emph{string} over $\Sigma$ is a sequence $w = a_1\ldots a_n$ of characters from $\Sigma$, with $\epsilon$ denoting the \emph{empty string}. 
We use $w_1\cdot w_2$ to denote the \emph{concatenation} of strings $w_1$ and $w_2$.
$\Sigma^*$ is the set of all strings over $\Sigma$, $\Sigma^+ = \Sigma^*\setminus \{\epsilon\}$, and 
A \emph{language} over $\Sigma$ is a subset $L$ of $\Sigma^*$.  
%
We use $|w|$ to denote the lenght of $w$ and $|w|_a$ to denote the number of occurrences of the character $a\in \Sigma$ in $w$.  
The \emph{Parikh image} of $w$ is the assignment 
%$\parikh(w) = \{a_1 \mapsto |w|_{a_1},\ldots, a_n \mapsto |w|_{a_n}\}$ that maps every symbol in $w$ to the number of its occurences in $w$. 
$\parikhof w$ that maps every character $a\in\Sigma$ to the number of its occurences in $w$, i.e., $\parikhof w (a) = |w|_a$. 
The Parikh image of a language $L$ is the set of Parikh images $\parikhof L = \{\parikhof w \mid w \in  L\}$.

@@

We introduce \emph{symbolic flat automata} (SFA) that will be used to define patterns 
%for the string variable domain restriction. 
used by the underapproximation module to restrict the domain of string variables. 
Namely, an SFA is a finite automaton $A = (Q,T,\cvar,\st{0}{0},\st{m}{0})$ operating over the alphabet $\cvar$ of \emph{character variables}, 
and satisfying the following structural constraints:
\begin{enumerate}
	\item The final state $\st{m}{0}$ is reached from  the initial state $\st{0}{0}$ through a stright path of $m-1$ transitions $(\st{i}{0},v_i,\st{i+1}{0}) \in T$, $\st{i}{0} \in Q$ for $i\in[0,m-1]$. %, with pairwise distiand states $\st{i}{0} \in Q$ for $i\in[0,m-1]$.
	\item  
Each state $\st{i}{0}$ is the origin of a unique simple cycle of the length $l_i\in\mathbb{N}$, consisting of states $\st{i}{j} \in Q$ and transitions $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$ for $j\in [1,l_i]$. 
Notice that case when $l_i = 0$ is also admissible and means that there is no cycle on $q_i$.
%Assume that $\st{i}{0}$ is included in a cycle with $l_i$ states (includes $\st{i}{0}$). 
%For each $j\in [1,l_i]$, we have $\st{i}{j} \in Q$ and $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$. 
%(where $l_i = 0$ means that there is no cycle on $q_i$)
	\item Each variable in $\cvar$ appears on at most one transition (but $\epsilon$ transition can appear multiple times). 
\end{enumerate} 
%Such automata are recognizers of symbolic strings, i.e., strings over $\cvar$.
Symbolic automata accept strings over $\cvar$, called \emph{symbolic strings}. 
Each symbolic string may be interpeted as a string over $\Sigma$, given an assignment of characters to its character variables.  
%
Formally, an \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$ and for a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

@@

%\section{Handling Equality and Disequality Constraint Symbolically} \label{section:eq}
\section{Flat Underapproximation for Equalities and Regular Constraints} \label{section:eq}

@@

Disquality constraint $x_1\cdot x_2 \cdots x_n \neq x_{n+1}\cdot x_{n+2} \cdots x_m$ can be converted to equal-satisfiable equality constraints in the standard way~\cite{abdulla2015norn}. We replace it with the following constraints.

@@

In this section, we discuss how to handle the constraint $n=\sti{x}$ by restrict the domain of $x$ to some SFA. Let us begin with a simple example, assume that we use the SFA in Figure~\ref{fig:sfa} (a) to restrict the domain of $x$. Then we know that when $0\leq v_0,v_1 \leq 9$, then $n$ is a positive integer value, and otherwise $n =-1$. So we should first add the constraint $ ((0\leq v_0\leq 9) \wedge (0\leq v_1\leq 9)) \vee n=-1 $.

For the case that $n$ is a positive integer, the value of $n$ can be characterized by a constraint $n= (v_0\times 10+ v_1) \times (1+100 +100^2 + \ldots 100 ^{\#(v_0)-1})=  (v_0\times 10+ v_1) \times \frac{100^{\#(v_0)+1}-1}{100-1}$. 
The constraint uses $\#(v_0)$ to capture the total number of loop traversal.
Notice that the constraint above contains an exponential component $100^{\#(v_0)}$. To solve the satisfiability of this formula, one need to solve an exponential constraint.

For a more general form, if we restrict the domain of $x$ to the SFA in Figure~\ref{fig:sfa} (c), for the case when $n$ is positive, we have the relation $n= (v_0^{(0)}\times 10+ v_1^{(0)}) \times (1+100 +100^2 + \ldots 100 ^{v_0^{(0)}-1})\times 100^{v_2^{(0)}}+(v_2^{(0)}\times 10+ v_3^{(0)}) \times (1+100 +100^2 + \ldots 100 ^{v_2^{(0)}-1}) =  (v_0^{(0)}\times 10+ v_1^{(0)}) \times \frac{100^{\#(v_0^{(0)})+1}-1}{100-1}\times 100^{v_2^{(0)}} + (v_2^{(0)}\times 10+ v_3^{(0)}) \times \frac{100^{\#(v_2^{(0)})+1}-1}{100-1}$. Observe that the formula has even more exponential components.

To the best of our knowledge, the satisfiability problem of integer constraints with a mix of polynomials and exponentials is still open.

For the particular case with only one integer variable, the satisfiability of integer constraints with a mix of polynomials and exponentials can be decided using the algorithm in~\cite{POPL2019}. However, the algorithm involves a quantifier elimination procedure, which is double-exponential to the length of the input formula and hence cannot handle large instances.

In order to handle the under-approximation of string-integer conversion efficiently, we propose to restrict the domain of $x$ to a SFA in Figure~\ref{fig:sfa_its}.

@@

$\begin{array}{ccc}
	(|x|&=& \#(v_0)+1 \wedge 0\leq v_1 \leq 9 \wedge v_2 = -1)\wedge \\
	(|x|&=& \#(v_0)+2 \wedge 0\leq v_2 \leq 9 \wedge v_3 = -1)\wedge \\
	& &\ldots\\
	(|x|&=& \#(v_0)+(m-1) \wedge 0\leq v_{m-1} \leq 9 \wedge v_m = -1)\wedge\\
	(|x|&=& \#(v_0)+m \wedge 0\leq v_m \leq 9 )
\end{array}$

@@

\section{Evaluation}
 \label{section:evaluation}
\todo{copy a few tables from our test site to here}
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

>>>>

\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\sti}[1]{\mbox{\textsf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textsf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{\mbox{$V^1_{\Sigma_\epsilon}$}}
\newcommand{\cvartwo}{\mbox{$V^2_{\Sigma_\epsilon}$}}
\newcommand{\modelof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}
\newcommand{\parikhof}[1]{\mathit{Parikh}{(#1)}}
\newcommand{\parikhwof}[2]{|#1|_{#2}}

@@

We use $\mathbb{N}$ and $\mathbb{Z}$ to denote the sets of natural numbers (including 0) and 
integers. For a set $A$, we use $|A|$ to denote its size. 
An \emph{alphabet} is a finite set $\Sigma$ of \emph{characters} and a \emph{string} over $\Sigma$ is a sequence $w = a_1\ldots a_n$ of characters from $\Sigma$, with $\epsilon$ denoting the \emph{empty string}. 
We use $w_1\cdot w_2$ to denote the \emph{concatenation} of strings $w_1$ and $w_2$.
$\Sigma^*$ is the set of all strings over $\Sigma$, $\Sigma^+ = \Sigma^*\setminus \{\epsilon\}$, and 
A \emph{language} over $\Sigma$ is a subset $L$ of $\Sigma^*$.  
%
We use $|w|$ to denote the length of $w$ and $|w|_a$ to denote the number of occurrences of the character $a\in \Sigma$ in $w$.  
The \emph{Parikh image} of $w$ is the assignment 
%$\parikh(w) = \{a_1 \mapsto |w|_{a_1},\ldots, a_n \mapsto |w|_{a_n}\}$ that maps every symbol in $w$ to the number of its occurences in $w$. 
$\parikhof w$ that maps every character $a\in\Sigma$ to the number of its occurrences in $w$, i.e., $\parikhof w (a) = |w|_a$. 
The Parikh image of a language $L$ is the set of Parikh images $\parikhof L = \{\parikhof w \mid w \in  L\}$.

@@

We introduce \emph{symbolic flat automata} (SFA) that will be used to define patterns 
%for the string variable domain restriction. 
used by the under-approximation module to restrict the domain of string variables. 
Namely, an SFA is a finite automaton $A = (Q,T,\cvar,\st{0}{0},\st{m}{0})$ operating over the alphabet $\cvar$ of \emph{character variables}, 
and satisfying the following structural constraints:
\begin{enumerate}
	\item The final state $\st{m}{0}$ is reached from  the initial state $\st{0}{0}$ through a stright path of $m-1$ transitions $(\st{i}{0},v_i,\st{i+1}{0}) \in T$, $\st{i}{0} \in Q$ for $i\in[0,m-1]$. %, with pairwise distiand states $\st{i}{0} \in Q$ for $i\in[0,m-1]$.
	\item  
Each state $\st{i}{0}$ is the origin of a unique simple cycle of the length $l_i\in\mathbb{N}$, consisting of states $\st{i}{j-1} \in Q$ and transitions $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$ for $j\in [1,l_i]$. 
Notice that case when $l_i = 0$ is also admissible and means that there is no cycle on $q_i$.
%Assume that $\st{i}{0}$ is included in a cycle with $l_i$ states (includes $\st{i}{0}$). 
%For each $j\in [1,l_i]$, we have $\st{i}{j} \in Q$ and $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$. 
%(where $l_i = 0$ means that there is no cycle on $q_i$)
	\item Each variable in $\cvar$ appears on at most one transition (but $\epsilon$ transition can appear multiple times). 
\end{enumerate} 
%Such automata are recognizers of symbolic strings, i.e., strings over $\cvar$.
Symbolic automata accept strings over $\cvar$, called \emph{symbolic strings}. 
Each symbolic string may be interpreted as a string over $\Sigma$, given an assignment of characters to its character variables.  
%
Formally, an \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$ and for a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

@@

%\section{Handling Equality and Disequality Constraint Symbolically} \label{section:eq}
\section{Flat Under-approximation for Equalities and Regular Constraints} \label{section:eq}

@@

Disequality constraint $x_1\cdot x_2 \cdots x_n \neq x_{n+1}\cdot x_{n+2} \cdots x_m$ can be converted to equal-satisfiable equality constraints in the standard way~\cite{abdulla2015norn}. We replace it with the following constraints.

@@

In this section, we discuss how to handle the constraint $n=\sti{x}$ by restricting the domain of $x$ to some SFA. Let us begin with a simple example, assume that we use the SFA in Figure~\ref{fig:sfa} (a) to restrict the domain of $x$. Then we know that when $0\leq v_0,v_1 \leq 9$, then $n$ is a positive integer value, and otherwise $n =-1$. So we should first add the constraint $ ((0\leq v_0\leq 9) \wedge (0\leq v_1\leq 9)) \vee n=-1 $.

For the case that $n$ is a positive integer, the value of $n$ can be characterized by a constraint $n= (v_0\times 10+ v_1) \times (1+100 +100^2 + \ldots 100 ^{\#(v_0)-1})=  (v_0\times 10+ v_1) \times \frac{100^{\#(v_0)}-1}{100-1}$. 
The constraint uses $\#(v_0)$ to capture the total number of loop traversal.
Notice that the constraint above contains an exponential component $\frac{100^{\#(v_0)}}{100-1}$. To solve the satisfiability of this formula, one need to solve an exponential constraint.

Let us have a look at another example. If we restrict the domain of $x$ to the SFA in Figure~\ref{fig:sfa} (c), for the case when $n$ is positive, we have the relation $n= (v_0^{(0)}\times 10+ v_1^{(0)}) \times (1+100 +100^2 + \ldots 100 ^{\#(v_0^{(0)})-1})\times 100^{\#(v_2^{(0)})}+(v_2^{(0)}\times 10+ v_3^{(0)}) \times (1+100 +100^2 + \ldots 100 ^{\#(v_2^{(0)})-1}) =  (v_0^{(0)}\times 10+ v_1^{(0)}) \times \frac{100^{\#(v_0^{(0)})}-1}{100-1}\times 100^{\#(v_2^{(0)})} + (v_2^{(0)}\times 10+ v_3^{(0)}) \times \frac{100^{\#(v_2^{(0)})}-1}{100-1}$. Observe that the formula has two exponential components $\frac{100^{\#(v_0^{(0)})}\times 100^{\#(v_2^{(0)})} }{100-1}$ and $\frac{100^{\#(v_2^{(0)})} }{100-1}$.

It is not difficult to observe that, for the case of an arbitrary SFA of $m$ loops, the formula to define $n$ contains $m$ exponential components, one for each loop.
To the best of our knowledge, the satisfiability problem of integer constraints with a mix of polynomials and exponentials is still open. As said in the introduction, the problem is difficult even for the case that variables are real numbers. For example, the algorithm in~\cite{kincaid2019closed} involves a quantifier elimination procedure, which is double-exponential to the length of the input formula and hence cannot handle large instances. So we do not expect that this constraint can be solved efficiently.

\todo{talk about the design principle}
In order to handle the under-approximation of string-integer conversion efficiently, we propose to restrict the domain of $x$ to a SFA in Figure~\ref{fig:sfa_its}.

@@

$\begin{array}{ccc}
	(|x|&=& \#(v_0)+1 \wedge 0\leq v_1 \leq 9 \wedge v_2 = -1)\vee \\
	(|x|&=& \#(v_0)+2 \wedge 0\leq v_2 \leq 9 \wedge v_3 = -1)\vee \\
	& &\ldots\\
	(|x|&=& \#(v_0)+(m-1) \wedge 0\leq v_{m-1} \leq 9 \wedge v_m = -1)\vee\\
	(|x|&=& \#(v_0)+m \wedge 0\leq v_m \leq 9 )
\end{array}$

@@

% \section{Evaluation}
%  \label{section:evaluation}
% \todo{copy a few tables from our test site to here}
\input{evaluation}

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}
